# Visualization

## Outline / cheat sheet  
* Plot architecture
  * `ggplot`: Create a blank canvas to plot on -- this is the foundation of your plot
  * `aes`: Identify variables in the data frame you'd like to map, including x and y coordinates as well as features like shape, size, color, etc.
  * `geom_`: Select from various geometries to give shape to your aesthetic mappings,
* Geometries
  * `geom_point`: Points for scatter plots, useful for continuous x continuous variables
  * `geom_smooth`: Add trend lines to a plot (especially scatter plots) to detect trends based on a specified estimation  
  * `geom_bar`: Bar graph for counting up observations of categorical variables
  * `geom_boxplot`: Box plots, useful for distribution of continuous x categorical
  * `geom_line`: Line plot, useful for tracking continuous variables over time
* Styling (automatic and manual)
  * `scale_`: Adjust any mapping (e.g. `scale_color_`, `scale_size_`) for customization  
  * `labs`: Add labels to any mapping
  * `theme_`: Adjust the overall look of plot (e.g. background, axis lines)  

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(palmerpenguins)
library(lubridate)
weather <- read.csv("https://pasta.lternet.edu/package/data/eml/knb-lter-pal/28/8/375b34051b162d84516ec2d02f864675") 
weather <- weather %>% 
  mutate(Date = ymd(Date),
         year = year(Date))
```

## Plot architecture

`ggplot` is an incredibly powerful plotting tool that follows the 'grammar of graphics' -- an approach that supports data visualization according to best practices. A `ggplot` has a template with three basic parts 1) data, 2) mappings, and 3) geometries, as well as countless other add-ons to enhance the plot. The generic syntax for combining these parts is below:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

We'll get into the add-ons throughout this lesson, but let's start by building a plot from the base up. First, let's input the penguins data into the ggplot function:  

```{r}
ggplot(data = penguins)
```

What we get is a blank canvas of a plot -- ggplot knows you want to plot something, but it isn't sure what yet because you haven't given it any specifics about the data. We get into the specifics with mappings, where we give the plot data-based features like x and y coordinates: 

```{r}
ggplot(data = penguins, mapping = aes(x = bill_length_mm,
                                      y = bill_depth_mm))
```

Now ggplot knows what you want to plot, and it maps them onto the x and y axes, as we specified. But we still haven't told ggplot _how_ to map them (i.e. in what shape?). We can specify the shape using one of several `geom_` functions. Let's select a 'point' geometry, which we use for a scatter plot of continuous x continuous variables.

```{r}
ggplot(data = penguins, mapping = aes(x = bill_length_mm,
                                      y = bill_depth_mm)) +
  geom_point()
```

Now we're getting somewhere. Let's play around a bit with some additional arguments that you can add to the plot, such as changing the transparency, size, shape, and color. These are all their own arguments that can be placed inside either the ggplot function or the geom function. I prefer to put them in the goem function.

```{r}
ggplot(data = penguins,
                         aes(x = bill_length_mm,
                             y = bill_depth_mm)) +
  geom_point(alpha = .8, size = 3, shape = 15, color = "darkorange")
```

What's powerful about ggplot is that you can add on multiple layers using the plus sign, and can even add several geometries to one plot. One geometry used for discerning trends is the `geom_smooth()` geometry, which will assign a trend line based on the distribution you specify, such as "lm", "glm", "gam", or "loess". Let's stick with a basic 'lm' method, which will create a trend line using the coefficient from the equation `lm(y ~ x)`, which in our case is `lm(bill_depth_mm ~ bill_length_mm, data = penguins)`

```{r}
ggplot(data = penguins,
                         aes(x = bill_length_mm,
                             y = bill_depth_mm)) +
  geom_point(alpha = .8, size = 3, shape = 15, color = "darkorange") +
  geom_smooth(method = "lm")
```

Here we see a slightly negative relationship between bill depth and length, which we can verify by running the same linear model we just passively applied using geom_smooth():

```{r}
lm(bill_depth_mm ~ bill_length_mm, data = penguins)
```

But remember, one of the reasons we visualize data is because it can be an incredibly helpful tool for exploratory data analysis, particularly because it gives you the opportunity to 'get to know' your data. If you are a penguin expert, you might be surprised by this result: bills tend to grow proportionally, right? And so we'd expect a positive relationship between these two variables. So, what else do we know about our data that might be hiding this relationship? Different penguin species! 

Species is an attribute of our data that we want to represent in our current plot, let's say with color. Previously, we used the color argument to add in a color, and you might want to follow that same pattern now.

```{r, eval = F}
ggplot(data = penguins, aes(x = bill_length_mm,
                            y = bill_depth_mm)) +
  geom_point(alpha = .8, size = 3, shape = 15, color = species) +
  geom_smooth(method = "lm", se = FALSE)
```

However, this approach won't work because ggplot doesn't know what "species" is, based on how we've input it into the code. We need a way to tell R that "species" is a column inside our penguins data. To do that, we include species as part of our aesthetic mapping, because the `aes()` function is what cues ggplot to look inside the data for values.

```{r}
ggplot(data = penguins,
               aes(x = bill_length_mm,
                   y = bill_depth_mm,
                   color = species)) +
  geom_point(alpha = .8, size = 3, shape = 15) +
  geom_smooth(method = "lm", se = FALSE)
```

Now its got it! So remember, if you want to add a feature to your plot that **is not related** to the data (e.g. color = "blue") it goes _outside_ the `aes` function. If you want to add a feature to your plot that **is related** to the data (e.g. color = Species), goes _inside_ the `aes` function. 

Now that we've got that straight, let's draw attention to what happened to our trends: Simpson's paradox! (i.e. 'statistical phenomenon where an association between two variables in a population emerges, disappears or reverses when the population is divided into subpopulations') ([Stanford 2021](https://plato.stanford.edu/entries/paradox-simpson/#:~:text=Simpson's%20Paradox%20is%20a%20statistical,population%20is%20divided%20into%20subpopulations.)). 

## Selecting geometries

Let's move on to discuss different geometries and when to use them. We've already noted that scatter plots with `geom_point` are a good candidate for continuous x continuous variables. What about when we start working with categorical variables? Let's review a few different approaches:

**geom_bar**: This bar plot geometry is well suited when you want to count up the observations of one categorical variable. By default, it takes only one mapping argument, x, and then calculates the counts based on the grouping of x.

```{r}
ggplot(penguins, aes(x = species)) + geom_bar()
```

**geom_boxplot**: This box plot geometry is well suited when you want to show the distribution of a continuous variable based on categorical groupings. Just like geom_point, geom_boxplot takes a x and a y argument, but one should be continuous and one should be categorical.

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) + geom_boxplot()
```

**geom_line**: This line geometry is well suited when you want to show how a continuous variable changes over time, where time is a bit of a mix between a continuous and categorical variable. 

```{r}
weather %>% 
  ggplot(aes(x = Date, y = Temperature.Average..C.)) +
  geom_line()
```

### Combining wrangling & visualization 

Let's practice plotting using the weather data, and let's think about our Date column as a categorical variable, with each year being a category. This will be a bit tricky, however, because the year variable is currently classified as a numeric data type, which means ggplot will want to treat it as continuous. 

```{r}
class(weather$year)
```

Because of this, ggplot gets a bit confused...

```{r}
weather %>% 
  ggplot(aes(x = year, y = Temperature.Average..C.)) +
  geom_boxplot() 
```

So, there are two ways we can prompt ggplot to think of year as a category. One way is to coerce our year variable into a 'factor', which tells R that these numbers are actually categories.

```{r}
weather %>% 
  ggplot(aes(x = factor(year), y = Temperature.Average..C.)) +
  geom_boxplot() 
```

Another way is to assign are group aesthetic.

```{r}
weather %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year)) +
  geom_boxplot() 
```

Now, remember the importance of using visualization to _explore_ data. Does anything look off?

It jumps out to me that our first and last years are shaped different than the others. They have no low-temperature outliers and their distributions are much smaller. Let's use our wrangling skills to investigate.

Take a look at a summary of the lowest (min) year:
```{r}
weather %>% 
  filter(year == min(year)) %>% 
  select(Date) %>% 
  summary(Date)
```

It only starts in April, so we are missing some data for that year. What about the final year:

```{r}
weather %>% 
  filter(year == max(year)) %>% 
  select(Date) %>% 
  summary(Date)
```

In ends in March, so we're also missing data for that year. So let's filter those out and pipe that filtered data frame directly into our ggplot:

```{r}
weather %>% 
  filter(year != min(year) & year != max(year)) %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year)) +
  geom_boxplot() 
```

### Check in challenge

Let's practice combining data wrangling with plotting. First, use the `case_when` function to `mutate` a new column called 'penguin_data_year'. The condition we'd like to set is to identify the years for which we have data in the 'penguins' data (2007:2009) as TRUE, and all of the other years FALSE. Second, re-create the yearly temperature box plot we just made, but add a color aesthetic using the 'penguin_year_data' to define the color. The end result should be the box plot, but with the distributions of 2007-2009 a different color from the rest. You can do this is two steps, or bonus if you pipe it all together.

<details>
<summary>Check your answer</summary>
```{r}
weather <- weather %>% 
  mutate(penguin_data_year = case_when(
    year %in% 2007:2009 ~ T,
    T ~ F
  ) )

weather %>% 
  filter(year != min(year) & year != max(year)) %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year,
             color = penguin_data_year)) +
  geom_boxplot() 
```
</details>

## Styling

For this last section, we are going to add some additional stylistic features to our plots that make them more presentable. Because `ggplot` works in layers, you can actually save a version of the plot as a object, and then use that to layer on different features without re-typing everything. So let's do that. 
```{r}
base_plot <- weather %>% 
  filter(year != min(year) & year != max(year)) %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year,
             color = penguin_data_year)) +
  geom_boxplot() 
```

Now we can add layers to that object. One of the first things we might want to do is set colors using pre-set palettes. We can adjust all of the mappings, such as color, using functions that start with `scale_<MAPPING>_`. One great set of palettes are linked to the `viridis` package, which makes colorblind friendly gradients, and ggplot includes the following functionality:  

```{r}
base_plot +
  scale_color_viridis_d()
```

Here we choose `_d` because we want a palette for a discrete (d) variable, not continuous (c).

We can also set colors manually with `scale_color_manual`, where we provide a vector of colors to the values argument.  

```{r}
base_plot +
  scale_color_manual(values = c("darkorange","cyan4"))
```

We can also add labels for all of the mappings with the `labs` function. 

```{r}
base_plot +
  labs(x = "Year", y = "Average Temperature (C)", 
       title = "Daily average temperature range at Palmer Station (1985-2018)",
       color = "Penguin data")
```

Next, `theme_` can add an overall nice touch to adjust the overall look of plot (e.g. background, axis lines). I like `theme_minimal` but there are all kinds of themes. Try tab completing after you type the following

```
theme_
```

Let's apply one: 

```{r}
base_plot +
  theme_minimal()
```

You can also further customize the theme with a stand-alone `theme` function. Here we can add customization like removing the legend.

```{r}
base_plot +
  theme_minimal() +
  theme(legend.position = "none")
```

Also within theme you can change the text (and much more!)

```{r}
base_plot +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(family = "Avenir"))
```

Now, let's put it all together

```{r}
base_plot +
  scale_color_manual(values = c("darkorange","cyan4")) +
  labs(x = "Year", y = "Average Temperature (C)", 
       title = "Daily average temperature range at Palmer Station (1985-2018)",
       color = "Penguin data") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(family = "Avenir"))
```

