# Visualization

## Outline
* Plot architecture
  * `ggplot`
  * `aes`
  * `geom_`
* Selecting geometries
  * `geom_point`
  * `geom_bar` and `geom_col`
  * `geom_boxplot`
  * `geom_line`
* Styling (automatic and manual)
  * `labs` 
  * `theme_`
  * `scale_`
* Facets

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(palmerpenguins)
library(lubridate)
weather <- read.csv("https://pasta.lternet.edu/package/data/eml/knb-lter-pal/28/8/375b34051b162d84516ec2d02f864675") 
weather <- weather %>% 
  mutate(Date = ymd(Date),
         year = year(Date))
```

## Plot architecture

Create a scatterplot of bill length and body mass

```{r}  
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g)) + 
  geom_point()
```

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g, color = species)) + geom_point()
```

## Selecting geometries

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g, color = species)) + 
  geom_point() +
  geom_smooth(method = 'lm')
```
Create a barplot of species counts
```{r}
ggplot(penguins, aes(x = species)) + geom_bar()
```

```{r}
ggplot(penguins, aes(x = species, y = flipper_length_mm)) + geom_boxplot()
```


```{r}
weather %>% 
  ggplot(aes(x = Date, y = Temperature.Average..C.)) +
  geom_line()
```


```{r}
weather %>% 
  ungroup() %>% 
  filter(year != min(year) & year != max(year)) %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year)) +
  geom_boxplot() 
```

## Challenge
Conditional statements

```{r}
weather <- weather %>% 
  mutate(penguin_data_year = case_when(
    year %in% 2007:2009 ~ T,
    T ~ F
  ) )
```

```{r}
weather %>% 
  filter(year != min(year) & year != max(year)) %>% 
  ggplot(aes(x = year, y = Temperature.Average..C., group = year,
             color = penguin_data_year)) +
  geom_boxplot() 
```

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g)) + geom_point() + facet_wrap(~ species)
```    

## Styling (automatic and manual)

